---
title: "Atlantic Salmon Decline in Argyll Rivers: A Time Series Analysis"
author: "Robert Jackson"
date: "April 2026"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Background and Research Question

Atlantic salmon (*Salmo salar*) populations across Scotland have declined markedly over the past 70 years. In Argyll, this decline is particularly well documented and has been linked to a range of pressures including climate change, marine mortality, and critically — the disruption of freshwater migration routes by hydropower infrastructure.

The River Awe is the principal salmon river in mid-Argyll, draining Loch Awe and its tributary catchment including the Rivers Orchy and Etive. The Inverawe hydropower scheme, operated by Scottish and Southern Energy (SSE), has regulated flow in the lower River Awe since 1965. The dam and associated fish pass at the Pass of Brander represents a significant barrier to upstream salmon migration. A change in the regulated freshet (artificial flood pulse) regime in 2012 was introduced specifically to improve upstream migration conditions.

This analysis uses the Marine Directorate's long-term rod catch statistics (1952–2023) to examine trends in Atlantic salmon catches across Argyll rivers, with particular focus on the River Awe catchment. The central question is: **does the long-term catch data reflect the known pressures on Argyll salmon populations, and is there any detectable signal around key management interventions?**

Rod catch data is used as a proxy for salmon abundance. While it is an imperfect measure — catch depends on fishing effort, catch-and-release rates, and reporting compliance — it represents the longest consistent time series available for Scottish rivers at district level.

---

## Data Source

Data are sourced from the Scottish Government Marine Directorate's open data portal:

> Scottish Government Marine Directorate. (2024). *Salmon and Sea Trout fishery statistics: 1952 to 2023 season — reported catch by district and method.* DOI: 10.7489/12458-1. Available at: https://data.marine.gov.scot

The rod catch dataset covers 107 Salmon Fishery Districts across Scotland, reporting monthly catches of wild multi sea-winter (MSW) and one sea-winter (1SW) salmon by method (retained and released) from 1952 to 2023. Released catch data is available from 1994 onwards.

---

## Load Packages

```{r packages}
# install any missing packages before loading
required_packages <- c("fpp3", "tidyverse", "kableExtra", "patchwork")
installed <- rownames(installed.packages())
to_install <- required_packages[!required_packages %in% installed]
if (length(to_install) > 0) install.packages(to_install)

library(fpp3)        # tsibble, fable, feasts for time series
library(tidyverse)   # data wrangling and ggplot2
library(kableExtra)  # formatted tables
library(patchwork)   # combining multiple plots
```

---

## Read and Explore the Data

```{r read-data}
# data is read directly from the Marine Directorate open data portal
# this ensures the analysis is fully reproducible without a local file

url <- "https://data.marine.gov.scot/sites/default/files//SalmonAndSeaTroutRodsByDistrictAndMonth1952-2023.csv"

salmon_raw <- read_csv(url)

# check structure
glimpse(salmon_raw)
```

```{r explore}
# how many districts and years?
n_distinct(salmon_raw$District)
range(salmon_raw$Year)

# what regions are present?
unique(salmon_raw$Region)

# what methods are recorded?
unique(salmon_raw$Method)
```

```{r argyll-districts}
# identify the districts within the West Coast region
# this covers the Argyll area including the Awe, Add and Fyne catchments
salmon_raw %>%
  filter(Region == "West Coast") %>%
  distinct(District, Region) %>%
  arrange(District) %>%
  kbl(caption = "West Coast Salmon Fishery Districts in the dataset") %>%
  kable_styling(full_width = FALSE)
```

---

## Build the Tidy Dataset

```{r tidy-data}
# filter to the key Argyll districts
# note: Etive does not appear in the rod catch dataset and is excluded
# the three districts with data are Add, Awe and Fyne

argyll_districts <- c("Awe", "Add", "Fyne")

# combine retained and released catch - total catch is the best available
# measure of how many salmon anglers are encountering each year
# aggregate MSW and 1SW into total wild salmon - both life history stages matter
# sum to annual totals - monthly variation is not the focus of this analysis

argyll_annual <- salmon_raw %>%
  filter(District %in% argyll_districts) %>%
  mutate(total_wild = `Wild MSW number` + `Wild 1SW number`) %>%
  group_by(District, Year) %>%
  summarise(total_catch = sum(total_wild, na.rm = TRUE),
            .groups = "drop")

# check completeness - important before modelling
argyll_annual %>%
  group_by(District) %>%
  summarise(n_years    = n(),
            first_year = min(Year),
            last_year  = max(Year),
            missing    = sum(is.na(total_catch))) %>%
  kbl(caption = "Data completeness by Argyll district") %>%
  kable_styling(full_width = FALSE)
```

```{r tsibble}
# convert to tsibble
# index = Year (the time variable)
# key   = District (identifies each individual river series)

argyll_ts <- argyll_annual %>%
  as_tsibble(index = Year, key = District)

glimpse(argyll_ts)
```

---

## Exploratory Analysis

### Catch Trends Across Argyll Rivers

```{r multi-river-plot, fig.width=10, fig.height=6}
# plot all three rivers to compare long-run trends
# free_y scale allows each river to show its own range clearly
argyll_ts %>%
  autoplot(total_catch) +
  facet_wrap(~ District, scales = "free_y", ncol = 2) +
  labs(title = "Figure 1: Annual Wild Atlantic Salmon Rod Catch — Argyll Rivers",
       subtitle = "Total wild catch (MSW + 1SW, retained and released) 1952-2023\nSource: Scottish Government Marine Directorate",
       y = "Total rod catch (number of fish)",
       x = "Year") +
  theme_minimal() +
  theme(legend.position = "none")

# all three rivers show long-run decline, consistent with the national picture
# the Awe is the largest river and shows the highest absolute catches
# the Add and Fyne are smaller catchments with lower totals throughout
```

```{r summary-table}
# summary statistics by district
argyll_annual %>%
  group_by(District) %>%
  summarise(
    years       = n(),
    mean_catch  = round(mean(total_catch, na.rm = TRUE), 0),
    max_catch   = max(total_catch, na.rm = TRUE),
    min_catch   = min(total_catch, na.rm = TRUE),
    peak_year   = Year[which.max(total_catch)],
    recent_mean = round(mean(total_catch[Year >= 2010], na.rm = TRUE), 0)
  ) %>%
  kbl(caption = "Summary statistics: Argyll salmon rod catch by district",
      col.names = c("District", "Years", "Mean catch", "Max catch",
                    "Min catch", "Peak year", "Mean 2010-2023")) %>%
  kable_styling(full_width = FALSE)
```

### Focus: River Awe

The River Awe is selected for detailed modelling because it is the largest salmon river in Argyll, has the most complete long-term record, and is the river most directly affected by hydropower regulation. The Inverawe scheme has controlled flows in the lower Awe since 1965.

```{r awe-only}
# extract the Awe series for detailed analysis
# the District key is dropped so the series is a simple single-variable tsibble
# this avoids key mismatch errors in the forecast and autoplot functions

awe_ts <- argyll_ts %>%
  filter(District == "Awe") %>%
  select(Year, total_catch) %>%
  as_tsibble(index = Year)
```

```{r awe-plot, fig.width=10, fig.height=5}
# main series plot with key management events annotated
awe_ts %>%
  autoplot(total_catch, colour = "steelblue", linewidth = 0.8) +
  geom_vline(xintercept = 1965, linetype = "dashed",
             colour = "firebrick", alpha = 0.7) +
  annotate("text", x = 1966, y = 1800,
           label = "Inverawe hydro\noperational (1965)",
           hjust = 0, size = 3, colour = "firebrick") +
  geom_vline(xintercept = 2012, linetype = "dashed",
             colour = "darkgreen", alpha = 0.7) +
  annotate("text", x = 2013, y = 1800,
           label = "Freshet regime\nchange (2012)",
           hjust = 0, size = 3, colour = "darkgreen") +
  labs(title = "Figure 2: River Awe — Annual Wild Salmon Rod Catch, 1952-2023",
       subtitle = "Key hydropower management events annotated\nSource: Scottish Government Marine Directorate",
       y = "Total rod catch (number of fish)",
       x = "Year") +
  theme_minimal()

# the series shows several distinct phases:
# 1952-1965: pre-regulation period, catches variable with peaks above 1500 fish
# 1965-1990: following Inverawe construction, catches decline and rarely exceed 800
# 1990-2010: continued low catches with high interannual variability
# 2010-2023: some of the lowest catches on record despite the 2012 freshet change
# no seasonal component as this is annual data
```

### ACF Analysis

```{r acf-plot, fig.width=8, fig.height=4}
# ACF shows how correlated each year's catch is with previous years
awe_ts %>%
  ACF(total_catch, lag_max = 20) %>%
  autoplot() +
  labs(title = "Figure 3: ACF — River Awe Annual Salmon Catch",
       subtitle = "Positive autocorrelation at early lags confirms persistent structure in the series",
       y = "ACF") +
  theme_minimal()

# positive autocorrelation at early lags declining to around lag 10
# this confirms each year's catch is related to the previous few years
# however the autocorrelation is not as extreme as a strongly trended series
# the high interannual noise in salmon catches is also evident in the ACF
```

### Comparing Awe and Add

The River Add has no major hydropower infrastructure and acts as a reference river. Both rivers share the same marine environment so marine mortality affects them equally — differences in their freshwater catch trends are therefore more informative about in-river pressures.

```{r comparison-plot, fig.width=10, fig.height=5}
# index both rivers to their 1952 baseline
# this removes the effect of different river sizes for a fair proportional comparison
argyll_ts %>%
  filter(District %in% c("Awe", "Add")) %>%
  group_by(District) %>%
  mutate(index = total_catch / total_catch[Year == 1952] * 100) %>%
  ungroup() %>%
  ggplot(aes(x = Year, y = index, colour = District)) +
  geom_line(linewidth = 0.8) +
  geom_hline(yintercept = 100, linetype = "dotted", colour = "grey50") +
  geom_vline(xintercept = 1965, linetype = "dashed",
             colour = "firebrick", alpha = 0.5) +
  scale_colour_manual(values = c("Awe" = "steelblue", "Add" = "orange")) +
  labs(title = "Figure 4: Indexed Salmon Catch — River Awe vs River Add (1952 = 100)",
       subtitle = "Proportional comparison removes the effect of different river sizes",
       y = "Catch index (1952 = 100)",
       x = "Year",
       colour = "River") +
  theme_minimal()

# a steeper post-1965 decline in the Awe relative to the Add would be
# consistent with an additional hydropower-related pressure on the Awe
```

---

## Time Series Modelling

Two ETS models are fitted to the River Awe series. ETS stands for Error, Trend, Seasonality — each component can be additive (A), multiplicative (M), or absent (N).

- **ETS(A,N,N)** — additive error, no trend, no seasonality. Simple exponential smoothing. Forecasts a flat line from the most recently estimated level.
- **ETS(A,A,N)** — additive error, additive trend, no seasonality. Holt's linear method. Adds a slope component, allowing the forecast to project a directional line.

Given the visible downward trend in the series, ETS(A,A,N) might be expected to fit better. However, the high interannual variability in salmon catches may limit the model's ability to reliably estimate a trend, and AIC — which penalises models for using additional parameters — will determine which is formally preferred.

```{r fit-models}
# fit both models to the River Awe series
awe_fit <- awe_ts %>%
  model(
    `ETS(A,N,N)` = ETS(total_catch ~ error("A") + trend("N") + season("N")),
    `ETS(A,A,N)` = ETS(total_catch ~ error("A") + trend("A") + season("N"))
  )

# ETS(A,N,N) model report
# alpha controls how quickly the level responds to new data
# alpha = 0.335 means the model gives reasonable weight to both recent
# and older observations rather than discarding older data entirely
awe_fit %>%
  select(`ETS(A,N,N)`) %>%
  report()

# ETS(A,A,N) model report
# alpha = level smoothing, beta = trend smoothing
# a very small beta means the trend estimate changes very slowly over time
awe_fit %>%
  select(`ETS(A,A,N)`) %>%
  report()
```

```{r model-accuracy}
# training accuracy comparison
accuracy(awe_fit) %>%
  select(.model, RMSE, MAE, MAPE) %>%
  kbl(caption = "Model accuracy on training data: River Awe salmon catch",
      digits = 3) %>%
  kable_styling(full_width = FALSE)
```

### Fitted vs Actual

```{r fitted-plots, fig.width=10, fig.height=8}
p1 <- augment(awe_fit) %>%
  filter(.model == "ETS(A,N,N)") %>%
  ggplot(aes(x = Year)) +
  geom_line(aes(y = total_catch, colour = "Actual")) +
  geom_line(aes(y = .fitted, colour = "Fitted"), linetype = "dashed") +
  scale_colour_manual(values = c("Actual" = "steelblue", "Fitted" = "orange")) +
  labs(title = "ETS(A,N,N) - Fitted vs Actual",
       y = "Rod catch", x = "Year", colour = "") +
  theme_minimal()

p2 <- augment(awe_fit) %>%
  filter(.model == "ETS(A,A,N)") %>%
  ggplot(aes(x = Year)) +
  geom_line(aes(y = total_catch, colour = "Actual")) +
  geom_line(aes(y = .fitted, colour = "Fitted"), linetype = "dashed") +
  scale_colour_manual(values = c("Actual" = "steelblue", "Fitted" = "orange")) +
  labs(title = "ETS(A,A,N) - Fitted vs Actual",
       y = "Rod catch", x = "Year", colour = "") +
  theme_minimal()

# patchwork stacks the two plots vertically
(p1 / p2) +
  plot_annotation(title = "Figure 5: Fitted vs Actual - River Awe Salmon Catch",
                  subtitle = "ETS(A,N,N) top, ETS(A,A,N) bottom")
```

### Forecasts

```{r forecast-plots, fig.width=10, fig.height=8}
p3 <- awe_fit %>%
  select(`ETS(A,N,N)`) %>%
  forecast(h = 10) %>%
  autoplot(awe_ts) +
  labs(title = "ETS(A,N,N) - 10-year forecast",
       y = "Rod catch", x = "Year") +
  theme_minimal()

p4 <- awe_fit %>%
  select(`ETS(A,A,N)`) %>%
  forecast(h = 10) %>%
  autoplot(awe_ts) +
  labs(title = "ETS(A,A,N) - 10-year forecast",
       y = "Rod catch", x = "Year") +
  theme_minimal()

(p3 / p4) +
  plot_annotation(
    title = "Figure 6: 10-year Forecasts - River Awe Salmon Catch",
    subtitle = "ETS(A,N,N) top, ETS(A,A,N) bottom. Shading = 80% and 95% prediction intervals")
```

```{r forecast-comparison, fig.width=10, fig.height=5}
# overlay both forecasts for direct comparison
awe_fit %>%
  forecast(h = 10) %>%
  autoplot(awe_ts, level = NULL) +
  labs(title = "Figure 7: Forecast Comparison - ETS(A,N,N) vs ETS(A,A,N)",
       subtitle = "Point forecasts only. River Awe wild salmon rod catch.",
       y = "Rod catch (number of fish)",
       x = "Year",
       colour = "Model") +
  theme_minimal()

# ETS(A,N,N) projects a flat line at the current level
# ETS(A,A,N) continues a very slight downward drift
# the practical difference between the two forecasts is modest
```

---

## Results and Conclusions

```{r final-table}
accuracy(awe_fit) %>%
  select(.model, RMSE, MAE, MAPE) %>%
  kbl(caption = "Final model comparison: River Awe ETS models",
      digits = 3,
      col.names = c("Model", "RMSE", "MAE", "MAPE (%)")) %>%
  kable_styling(full_width = FALSE)
```

### Long-run Decline

The River Awe salmon catch has declined substantially over the 72-year record. Peak catches in the early post-war period exceeded 1,500 fish per year. By the 2010s, annual catches had fallen below 200 fish in most years — a decline of over 85% from peak. This is consistent with national trends: Atlantic salmon were reclassified by the IUCN Red List as Endangered in Great Britain in December 2023, reflecting a 30–50% decline in British populations since 2006.

### Hydropower Signal

The indexed comparison between the Awe and Add (Figure 4) shows that both rivers have declined over the long run, consistent with the broader national picture. However, the Awe shows a notably steeper decline in the period following the construction of the Inverawe scheme in 1965. The Add, which has no major in-river barriers, does not show the same sharp post-1965 trajectory. This is circumstantial rather than conclusive — a rigorous analysis would need to control for differences in fishing effort and other local factors — but the pattern is consistent with the hypothesis that hydropower regulation has imposed an additional pressure on the Awe above the background national decline.

### 2012 Freshet Intervention

The freshet regime change in 2012 does not produce a clear recovery signal in the catch data. Catches in the years following 2012 include some of the lowest on record. However, salmon population recovery is slow — any benefit from improved migration conditions would take multiple generations (at minimum 5–7 years) to become detectable in catch data, and multiple other pressures continue to operate simultaneously. It is not possible to conclude from this dataset alone whether the intervention has or has not been effective.

### Model Comparison

An interesting result emerges from the model comparison. ETS(A,A,N) has a marginally lower RMSE but a higher AIC (1118.00) than ETS(A,N,N) (AIC = 1114.99). AIC penalises models for additional parameters, so despite the slight improvement in training fit, the extra beta parameter in ETS(A,A,N) is not sufficiently justified by the data. The near-zero beta value (0.0001) confirms this — the model has set its trend component almost to zero, defaulting to near-identical behaviour to ETS(A,N,N). This happens because the high interannual variability in salmon catches makes it difficult to reliably separate a consistent trend from noise, even though the visual downward trajectory is clear across the full 72-year record.

On AIC grounds, ETS(A,N,N) is the formally preferred model. This is a useful reminder that statistical model selection is not always the same as ecological interpretation — the population decline is real and well documented, but a simple exponential smoothing model captures the data as well as a trend model given the level of noise in annual catch statistics.

### Limitations

Rod catch is an imperfect proxy for salmon abundance. Changes in fishing effort, the near-universal shift to catch-and-release since the 1990s, and variation in reporting compliance all affect the record. Fish counter data from the Pass of Brander — which directly counts fish passing the hydropower dam — would provide a more reliable measure of upstream migration success and would be the logical next step for this analysis. A mixed-effects model incorporating river-level covariates such as flow regime, water temperature, and sea lice burden would allow more formal separation of the different pressures acting on these populations.

---

## Data Citation

Scottish Government Marine Directorate. (2024). *Salmon and Sea Trout fishery statistics: 1952 to 2023 season — reported catch by district and method.* Available at: https://data.marine.gov.scot/dataset/salmon-and-sea-trout-fishery-statistics-1952-2023-season-reported-catch-district-and-method

---
*Analysis completed April 2026. All data read directly from the Marine Directorate open data portal. File knits to HTML without requiring any local data files.*
